A* with Fast and Tight Heuristic in Road Networks

### 2020-08-12 paper/ch_potentials.tex
ev routing citation
evaluation, finally

\usepackage{amssymb}
\usepackage{amsmath}

@@

\section{Introduction}
\label{sec:intro}
Route planning in street networks of continental scale is a well-researched topic with many applications~\cite{bdgmpsww-rptn-16}.
It is usually formalized as a variant of the classical shortest path problem in weighted graphs.
Efficiently computing flexible and exact routes is a thought challenge.
\emph{Exact} means that the computed path always has minimum weight.
\emph{Efficiently} means that the path computation is fast.
Being able to change the edge weights between path computations is \emph{flexibility}.

@@

There is a lot of work that extends CH to handle more complex settings.
For example, in \cite{fns-opca-14,gks-rpfof-10} multi-criteria optimization is studied.
In \cite{dsw-cch-15}, CH is modified to support Live-Traffic.
Historic traffic patterns are combined with CH in \cite{swz-sfert-19,bgsv-mtdtt-13,bdpw-dtdrp-16}.
Specialized variants for electric vehicle routing are developped in \cite{bdgwz-sfpcs-19}.
While these works show that it is possible to extend hierarchical approaches, they also show that it is non-trivial.
Further, in every setting the flexibility available at query time is fairly limited and usually encompasses only one very specific setting.
Combining all these hierarchical extensions is an unsolved problem.
%
CH-Potentials is not the first work to combine hierarchical approaches and A* \cite{bdsssw-chgds-10,gkw-blwr-07,bdgwz-sfpcs-19}.
However, past work mostly uses A* to accelerate hierarchical approaches even further and sacrifzies A*'s felxibility in the process.
Our primary algorithm design objective is to retain A*'s flexibility.

@@

Many applications need flexibility.
In road routing, the most common ones are traffic avoidance, user preferences, and vehicle restrictions.
Traffic avoidance comes in two forms: Live~\cite{CRP,CCH} and predicted traffic~\cite{td-alt,td-ch}.
Predicted traffic is often called time-dependent routing.
%
User preferences can vary significantly~\cite{DBLP:conf/gis/FunkeS15,DBLP:conf/gis/DellingGGKTW15,DBLP:conf/gis/FunkeLS16}
For example, for some users shorter routes are more important fast routes.
Others want to avoid tunnels or highways.
%
Truck routing differs from car routing because, depending on their load, trucks are not allowed everywhere~\cite{kswz-erptd-20}.
%
Flexibility is also often used as building block in multi-agent planning settings~\cite{DBLP:journals/ai/SharonSFS15,DBLP:journals/tciaig/BaierBHH15}.

@@

Formally, we set $w_q(e,\tau)$ to $w_c(e)$, if $\tau_{\mathrm{soon}}$.
Otherwise, we check whether $w_p(e,\tau_{\mathrm{soon}}) < w_c(e)$ is true.
If it is the case, we set $w_q(e,\tau)$ to $\max\{w_c(e)+(\tau_{\mathrm{soon}}-\tau), w_p(e,\tau)\}$.
Otherwise, we set $w_q(e,\tau)$ to $\min\{w_c(e)-(\tau_{\mathrm{soon}}-\tau), w_p(e,\tau)\}$.
We set $w_\ell$ again to the freeflow travel time.
Finally, we run the modified time-dependent A* described in the previous section on this modified travel time function. 
In our implementation, we to not modify the representation of $w_p$.
Instead, we evaluate the formulas above each time that the modified travel time function is evaluated.

@@

\section{Experiments}

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{fig/scaled_weights.pdf}
\caption{Query running times for queries of DIMACs Europe with scaled edge weights.}\label{fig:scaled_weights}
\end{figure*}

In this section, we present our experimental evaluation.
Our benchmark machine runs openSUSE Leap 15.1 (kernel 4.12.14), and has 128\,GiB of DDR4-2133 RAM and an Intel Xeon E5-1630 v3 CPUs which has four cores clocked at 3.7\,Ghz and 4~$\times$~32\,KiB of L1, 8~$\times$~256\,KiB of L2, and 10\,MiB of shared L3 cache.
All experiments were performed sequentially.
Our code is written in Rust and compiled with rustc 1.44.0-nightly in the release profile with the target-cpu=native option.

\subparagraph{Inputs and Methodology}
Our main benchmark instance is a graph of the road network of Western Europe made publicly available for the Ninth DIMACS implementation Challenge~\cite{DemetrescuGJ09} with 18M nodes and 42M edges.
We use this graph with the travel time as the lower bound and perform experiments with modified edge weights (scaling all travel times or reproducing an experiment from~\cite{dw-lbrdg-07} where 1\,000 random edges are perturbed).
Additionally, we perform experiments on a recent OSM instance of Germany\footnote{\url{https://download.geofabrik.de/europe/germany.html}, obtained on August 2nd, 2019}.
To obtain the routing graph, we use the import from RoutingKit\footnote{\url{https://github.com/RoutingKit/RoutingKit}}.
The graph has 11M nodes and 26M edges.
For this instance, we have proprietary live traffic snapshot from Friday 2019/08/02 afternoon provided by Mapbox\footnote{\url{https://www.mapbox.com/}}.
The live data comes in the form of 320K OSM node pairs and live speeds for the link between the nodes.
We also have two graphs with traffic predictions provided by PTV\footnote{\url{https://ptvgroup.com}}.
One is an old instance of Germany with traffic predictions from 2006 (4.7M nodes, 10.7M edges) and the other a newer instance of Europe (25.8M nodes, 55.5M edges, from 2017).
Preprocessing running times are averages over 10 runs.
For the queries, we perform 10\,000 point-to-point queries where both source and target are nodes drawn uniformly at random and report average results.

\begin{table*}
\centering
\caption{Query performance with the different algorithmic features.}\label{tab:building_blocks}
\input{table/building_blocks.tex}
\end{table*}

\begin{table*}
\centering
% TODO dijkstra baseline, preprocessing
\caption{Performance for different applications.}\label{tab:applications}
\input{table/applications.tex}
\end{table*}

@@

TODO

>>>>

\usepackage{amssymb}
\usepackage{amsmath}
% checkmarks
\usepackage{pifont}
\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}

@@

\section{Introduction}
\label{sec:intro}
Route planning in street networks of continental scale is a well-researched topic with many applications~\cite{bdgmpsww-rptn-16}.
It is usually formalized as a variant of the classical shortest path problem in weighted graphs.
Efficiently computing flexible and exact routes is a though challenge.
\emph{Exact} means that the computed path always has minimum weight.
\emph{Efficiently} means that the path computation is fast.
Being able to change the edge weights between path computations is \emph{flexibility}.

@@

There is a lot of work that extends CH to handle more complex settings.
For example, in \cite{fns-opca-14,gks-rpfof-10} multi-criteria optimization is studied.
In \cite{dsw-cch-15}, CH is modified to support Live-Traffic.
Historic traffic patterns are combined with CH in \cite{swz-sfert-19,bgsv-mtdtt-13,bdpw-dtdrp-16}.
Specialized variants for electric vehicle routing are developped in \cite{bdgwz-sfpcs-19}.
While these works show that it is possible to extend hierarchical approaches, they also show that it is non-trivial.
Further, in every setting the flexibility available at query time is fairly limited and usually encompasses only one very specific setting.
Combining all these hierarchical extensions is an unsolved problem.
%
CH-Potentials is not the first work to combine hierarchical approaches and A* \cite{bdsssw-chgds-10,gkw-blwr-07,bdgwz-sfpcs-19}.
However, past work mostly uses A* to accelerate hierarchical approaches even further and sacrifices A*'s flexibility in the process.
Our primary algorithm design objective is to retain A*'s flexibility.

@@

Many applications need flexibility.
In road routing, the most common ones are traffic avoidance, user preferences, and vehicle restrictions.
Traffic avoidance comes in two forms: Live~\cite{CRP,CCH} and predicted traffic~\cite{td-alt,td-ch}.
Predicted traffic is often called time-dependent routing.
%
User preferences can vary significantly~\cite{DBLP:conf/gis/FunkeS15,DBLP:conf/gis/DellingGGKTW15,DBLP:conf/gis/FunkeLS16}
For example, for some users shorter routes are more important fast routes.
Others want to avoid tunnels or highways.
%
Truck routing differs from car routing because, depending on their load, trucks are not allowed everywhere~\cite{kswz-erptd-20}.
%
Flexibility is also often used as building block in multi-agent planning settings~\cite{DBLP:journals/ai/SharonSFS15,DBLP:journals/tciaig/BaierBHH15}.
%
Electric Vehicle Routing also requires flexibility~\cite{DBLP:journals/algorithmica/BaumDPSWZ20,DBLP:conf/aaai/EisnerFS11}.

@@

Formally, we set $w_q(e,\tau)$ to $w_c(e)$, if $\tau \leq \tau_{\mathrm{soon}}$.
Otherwise, we check whether $w_p(e,\tau_{\mathrm{soon}}) < w_c(e)$ is true.
If it is the case, we set $w_q(e,\tau)$ to $\max\{w_c(e)+(\tau_{\mathrm{soon}}-\tau), w_p(e,\tau)\}$.
Otherwise, we set $w_q(e,\tau)$ to $\min\{w_c(e)-(\tau_{\mathrm{soon}}-\tau), w_p(e,\tau)\}$.
We set $w_\ell$ again to the freeflow travel time.
Finally, we run the modified time-dependent A* described in the previous section on this modified travel time function.
In our implementation, we to not modify the representation of $w_p$.
Instead, we evaluate the formulas above each time that the modified travel time function is evaluated.

@@

\section{Evaluation}

@@

\begin{table}
\centering
\caption{Instances used in the evaluation.}\label{tab:graphs}
\input{table/graphs.tex}
\end{table}

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{fig/scaled_weights.pdf}
\caption{
Running times on a logarithmic scale for queries on OSM Ger with scaled edge weights $w_q = \alpha \cdot w_\ell$.
The boxes cover the range between the first and third quartile.
The band in the box indicates the median, the diamond the mean.
The whiskers cover 1.5 times the inter quartile range.
All other running times are marked as outliers.
}\label{fig:scaled_weights}
\end{figure*}

In this section, we present our experimental evaluation.
Our benchmark machine runs openSUSE Leap 15.1 (kernel 4.12.14), and has 128\,GiB of DDR4-2133 RAM and an Intel Xeon E5-1630 v3 CPUs which has four cores clocked at 3.7\,Ghz and 4~$\times$~32\,KiB of L1, 8~$\times$~256\,KiB of L2, and 10\,MiB of shared L3 cache.
All experiments were performed sequentially.
Our code is written in Rust and compiled with rustc 1.47.0-nightly in the release profile with the target-cpu=native option.

\subparagraph{Inputs and Methodology}
Our main benchmark instance is a graph of the road network of Germany obtained from Open Street Map\footnote{\url{https://download.geofabrik.de/europe/germany-200101.osm.pbf}}.
To obtain the routing graph, we use the import from RoutingKit\footnote{\url{https://github.com/RoutingKit/RoutingKit}}.
The graph has 16M nodes and 35M edges.
For this instance, we have proprietary traffic data provided by Mapbox\footnote{\url{https://www.mapbox.com/}}.
The  data includes a live traffic snapshot from Friday 2019/08/02 afternoon and comes in the form of 320K OSM node pairs and live speeds for the link between the nodes.
It also includes traffic predictions for 38\% of the links as predicted speeds for all five minute periods over the course of a week.
Additionally, have two graphs with traffic predictions provided by PTV\footnote{\url{https://ptvgroup.com}}.
One is an old instance of Germany with traffic predictions from 2006 (4.7M nodes, 10.7M edges) and the other a newer instance of Europe (25.8M nodes, 55.5M edges, from 2017).
Table~\ref{tab:graphs} contains an overview over our instances.
We report preprocessing running times as averages over 10 runs.
For the queries, we perform 10\,000 point-to-point queries where both source and target are nodes drawn uniformly at random and report average results.

\subparagraph{Experiments}
We first investigate the effectiveness of the heuristic depending on how tight the lower bound is.
For that, we take our main graph, use the unmodified travel time metric as the lower bound $w_\ell$, and perform queries on $w_q = \alpha \cdot w_\ell$ where $\alpha$ is a factor $\geq 1$.
When $\alpha = 1$, the heuristic is as tight as possible.
With growing $\alpha$ the heuristic degrades.
Figure~\ref{fig:scaled_weights} depicts the results.
Clearly, $\alpha$ has significant influence on the running time.
Average running times range from below a millisecond to a few hundred milliseconds depending on $\alpha$.
Up to around $\alpha = 1.1$ the running time growth is roughly quadratic before leveling off to something more resembling linear growth.
We observe a significant amount of heavy outliers sometimes more than an order of magnitude slower than the median.
Generally, running times vary quite strongly even though most queries are long range queries as we draw source and target node uniformly at random.
The reason is that different queries (even with similar shortest path lengths) are affected differently by growing $\alpha$ depending on the geographic regions the shortest path traverses.
When $\alpha$ grows, more and more nodes in the area around the shortest paths are included in the search space.
So when the shortest path traverses an area where many nodes are reachable in short distance, the search space grows stronger, than when the shortest path goes only through sparse areas.
% Conclusions?

\begin{table}
\centering
\caption{Query performance with different heuristics and optimizations on OSM Ger with $w_q = 1.05 \cdot w_\ell$.}\label{tab:building_blocks}
\input{table/building_blocks.tex}
\end{table}

In Table~\ref{tab:building_blocks} we investigate the impact of our optimizations.
We run queries with different heuristic functions: the \emph{Zero} heuristic which turns A* back into the classical algorithm of Dijkstra, our new CH-Potential heuristic, and an \emph{Oracle} heuristic where all distances to $t$ with regards to $w_\ell$ can be obtained immediately.
We first observe that each one of our optimizations gains useful speedups regardless of the heuristic used.
Our optimizations do not only reduce heuristic evaluations but also queue operations which always accelerates the algorithm.
The number of queue pushs roughly corresponds to the running time for all heuristics.
When comparing CH-Potentials to the Oracle heuristic we note that CH-Potentials are only a factor of 1.6 slower.
This clearly shows that CH-Potentials have very little overhead.
Assuming that $w_\ell$ is the tightest possible lower bound that can be known in advance, no other heuristic can be more than a factor of 1.6 faster than CH-Potentials without sacrificing exactness.

\begin{table*}
\centering
\caption{
CH-Potentials performance for different route planning applications.
We report average running times and number of queue pushs.
We also report the average length increase, that is how much longer the final shortest distance is compared to the lower bound.
Finally, we report the average running time of Dijkstras algorithm as a baseline and the speedup of CH-Potentials over the baseline.
}\label{tab:applications}
\input{table/applications.tex}
\end{table*}

Table~\ref{tab:applications} depicts results when applying our algorithm to different route planning problems.
The greatest speedup is, of course, achieved when applying CH-Potentials to an unmodified graph with $w_q = w_\ell$.
In that case the performance is comparable to what other state-of-the-art route planning techniques like CH achieve.
When considering extended problem settings, the performance of CH-Potentials strongly depends on the tightness of $w_\ell$, as previously observed in Figure~\ref{fig:scaled_weights}.
On one hand, forbidding highways leads to over 40\% longer paths and the smallest speedup over Dijkstra (only 5.3) among all settings.
On the other hand, incorporating turn restrictions increase path lengths only marginally and CH-Potentials yield a speedup of around 1500 over Dijkstra on the turn expanded graph without any additional implementation complexity. % other techniques need crazy shit, cite turn papers
Live traffic increases shortest paths lengths by around 15\% longer which yields running times of 125\,ms on average.
The Mapbox traffic predictions increase shortest path lengths even stronger and running times increase correspondingly.
The PTV traffic predictions yield much smaller length increases (less than 4\%) which leads to much better running times and speedups.
Assessing which of these predictions is more realistic is not possible within the scope of this paper.
Combining Mapbox live traffic and traffic predictions yields running times only slightly slower than traffic predictions alone.
To the best of our knowledge, no other exact technique so far could combine these two scenarios.
Additionally adding turn restrictions slows running times down to around 850\,ms which is still a speedup of 7.3 over the baseline despite the length increase only growing by 1.5 percentage points.
The reason for this is that the turn expanded graph is three times bigger and consists of only one biconnected component, which disables our optimization from Section~\ref{sec:largested-biconnected-component}.
To the best of our knowledge no other approach so far did even attempt handling this combination of scenarios.
While the achieved speedup in our case is not particularly large, the reason is not the complexity of the scenario but the length increase in the time-dependent data.

@@

In this paper, we CH-Potentials introduced a fast and exact new A* heuristic for finding shortest paths in road networks.
The approach is flexible and can handle a multitude of routing scenarios with very small implementation complexity.
However, the performance of CH-Potentials depends on the tightness of the lower bound, so whether CH-Potentials are fit for a certain application depends on whether a tight lower bound can be derived.
Nevertheless, our experiments show that CH-Potentials are efficient in extracting distance estimates from a lower bound graph and only around 1.6 times slower than extracting lower bound distances from a precomputed matrix.
